---
title: 'Working from home and digital divides: resilience during the pandemic. Supplemental Material.'
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    orientation: columns
    social: menu
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../paper/v2_taylor_francis/")
  })
---

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(rmarkdown)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(openair)
library(rgdal)
library(maps)
library(rprojroot)
library(kableExtra)
#library(cowplot)
library(ggthemes)
library(leaflet)

path <- find_rstudio_root_file()
```

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
# load the data created by Data_Spatial.Rmd

path.data <- paste(path, "/data/temp/TSbb19_20sp.csv", sep = "")

TSbb19_20sp <- read_csv(path.data)

#split by year before create stats by each hour of each day of the week
TS2019 <- selectByDate(TSbb19_20sp, year = 2019) #272248 obs
TS2020 <- selectByDate(TSbb19_20sp, year = 2020) #343268 obs

# read clusters file
path.clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
clusters.nodiff <- read_csv(path.clusters)
```

Column {data-width=800}
-----------------------------------------------------------------------

### Upload speed clusters for Local Authority Districts, 2020. See [Budnitz and Tranos (2021)](https://doi.org/10.1080/24694452.2021.1939647) for more details. 

```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.asp=T}

# LAD fully generalised
# la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson",
#               verbose = FALSE)
# saved locally in /data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BUC.geojson

path.geojson <- paste0(path, "/data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BGC.geojson")
la <- readOGR(path.geojson, verbose = FALSE)

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))

la@data$lad19nm <- as.character(la@data$lad19nm)

# upload metrics
for.tip.mean.up.am <- TS2020 %>% # raw data, individual tests
  filter(hour(date)==9 | hour(date)==10) %>% 
  group_by(LAD19NM) %>%
  summarise(mean.up.am = mean(speedup)) %>% 
  mutate(rank = dense_rank(desc(mean.up.am)))

for.tip.mean.up.pm <- TS2020 %>% # raw data, individual tests
  filter(hour(date)==19 | hour(date)==20) %>%
  group_by(LAD19NM) %>%
  summarise(mean.up.pm = mean(speedup))

# not used
# for.tip.mean.up <- TS2020 %>% # raw data, individual tests
#   group_by(LAD19NM) %>%
#   summarise(mean.up = mean(speedup)) 
# 
# # download metrics
# for.tip.mean.down.am <- TS2020 %>% # raw data, individual tests
#   filter(hour(date)==9 | hour(date)==10) %>% 
#   group_by(LAD19NM) %>%
#   summarise(mean.down.am = mean(speeddown)) 
# 
# for.tip.mean.up.pm <- TS2020 %>% # raw data, individual tests
#   filter(hour(date)==19 | hour(date)==20) %>% 
#   group_by(LAD19NM) %>%
#   summarise(mean.down.pm = mean(speeddown)) 
# 
# for.tip.mean.up <- TS2020 %>% # raw data, individual tests
#   group_by(LAD19NM) %>%
#   summarise(mean.down = mean(speeddown)) 


# merge with the data
la <- sp::merge(la, clusters.nodiff, by.x = "lad19nm", by.y = "LAD")

# create the text to tooltip
la <- sp::merge(la, for.tip.mean.up.am, by.x = "lad19nm", by.y = "LAD19NM", all.x=T)
la <- sp::merge(la, for.tip.mean.up.pm, by.x = "lad19nm", by.y = "LAD19NM", all.x=T)

#sapply(la@data, function(x) sum(is.na(x)))
#No am data points for Isles of Scilly

# text for labels
la@data$tip <- paste0(la@data$lad19nm, "<br>",
                   "Mean upload AM speed: ", round(la@data$mean.up.am), " Kbps", "<br>",
                   "Mean upload PM speed: ", round(la@data$mean.up.pm), " Kbps") %>% 
  lapply(htmltools::HTML)

# using world.cities from `maps`
cities <- world.cities %>% 
  filter(country.etc=="UK") %>% 
  arrange(pop) %>% tail(10)

# Create a color palette for the map:
mypalette <- colorFactor(palette="Set1", 
                         domain=la$cluster.up, 
                         na.color="transparent",
                         reverse = T)

leaflet(la) %>% 
  addTiles()  %>% 
  fitBounds(-5.7, 50, -1.2, 58.7) %>% 
  addPolygons(fillColor = ~mypalette(cluster.up), 
              stroke=T,
              color = "white",
              weight = .8,
              opacity = .3,
              fillOpacity = .5,
              label = ~tip) %>% 
  addCircles(lng = cities$long, lat = cities$lat,
             weight = 2, radius = 5, color = "black",
             label = cities$name,
             labelOptions = labelOptions(noHide = T, direction = 'right', 
                                         textOnly = TRUE, permanent = F),
             options = markerOptions(riseOnHover = TRUE)) %>% 
  # addLabelOnlyMarkers(data = cities,
  #                   lng = ~long, lat = ~lat, label = ~name,
  #                   labelOptions = labelOptions(noHide = T, direction = 'auto', textOnly = TRUE)) %>%
  addProviderTiles(providers$Esri.WorldTerrain) %>% 
  addLegend(pal = mypalette,
            values = ~cluster.up,
            position = "topleft",
            opacity = .5,
            title = "Upload speed<br>clusters")

```

Column 
-----------------------------------------------------------------------

### Local Authority District mean AM and PM upload speeds

```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE,fig.height=10}
for.tip.mean.up.am <- merge(for.tip.mean.up.am, clusters.nodiff, 
                            by.x = "LAD19NM", by.y = "LAD", all.x = T)
for.tip.mean.up.am <- merge(for.tip.mean.up.am, for.tip.mean.up.pm, 
                            by.x = "LAD19NM", all.x = T)

for.tip.mean.up.am %>% 
  select(-cluster.down, -cluster.tests, -rank) %>% 
  arrange(-mean.up.am) %>% 
  relocate(cluster.up, .after=mean.up.pm) %>% 
  kable(digits = 0,
        format.args = list(big.mark = ","),
        col.names = c("LAD",
                      "Mean AM upload speed",
                      "Mean PM upload speed",
                      "Cluster")) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box() #width = "500px", height = "400px"
```

### Cluster description

- Cluster 1: Medium sized mostly rural cluster; slow mean upload speeds; high workday temporal variation.

- Cluster 2: Small rural cluster; highest mean upload speeds; slowdown more in evening.

- Cluster 3: Medium sized mixed cluster; fast mean upload speeds; medium temporal variation.

- Cluster 4: Rural reference LAD; medium mean upload speeds; high temporal variation.

- Cluster 5: Small suburban cluster; fast mean upload speeds; slowdown more in evening.

- Cluster 6: Large mixed cluster; slowest aggregate mean upload speeds; low workday temporal variation.

- Cluster 7: Small suburban cluster; fast mean upload speeds; medium temporal variation.

- Cluster 8: Small mainly surburban cluster; slow mean upload speeds; high workday temporal variation.

- Cluster 9: Large, more urban cluster; fast mean upload speeds; lowest workday temporal variation.
